---
title: "ST 558 Project 1"
author: "Carlos Petzold"
format: html
---

# Introduction

Here we are working on our first major project for ST 558, let's do it! Our overall goal is to write functions that will manipulate and process data that comes from the Census API, and write out our thought process as we work through it. 

The data that we are using will be queried from the Public Use Microdata Sample (PUMS) Census API, which gives person-level data. There are many different years included, but we will focus on 2021-2024 for this project, and we are also excluding Puerto Rico.


# Part 1: Data Processing

We'll begin by downloading and parsing the JSON files with the variable descriptions for each year:

```{r Part 1 - Parsing Variable Values}

#pull needed libraries
library(httr)
library(jsonlite)

#Using the Glue library for "better looking" paths
library(glue)

#setting desired variables for 2021-22 data
vars_21_22 <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION")

#setting desired variables for 2023-24 data
vars_23_24 <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION")

#Function for pulling census data by year and variables
pull_census_data <- function(year, vars) {
  
  #grabbing URL based on year 
  url <- glue("https://api.census.gov/data/{year}/acs/acs1/pums/variables.json")
  data <- GET(url)
  
  #Stop the function if the data request fails
  if (status_code(data) != 200) {
    stop(glue("Error, Census data request failed for year {year}: status {status_code(data)}"))
  }
  
  #Grab parsed content from the data we pulled
  parsed <- content(data, as = "text", encoding = "UTF-8") |> fromJSON()
  
  #Filter to our desired variables
  parsed$variables[vars]
  
}

#running our parsing function for years 2021-2024 to obtain one list for each year
vars_2021 <- pull_census_data(2021, vars_21_22)
vars_2022 <- pull_census_data(2022, vars_21_22)
vars_2023 <- pull_census_data(2023, vars_23_24)
vars_2024 <- pull_census_data(2024, vars_23_24)

#Saving our lists as an RDS file in our repo under a data folder
census_var_list <- list("2021" = vars_2021, "2022" = vars_2022, "2023" = vars_2023, "2024" = vars_2024)

dir.create("data", showWarnings = FALSE)
saveRDS(census_var_list, file = "data/census_var_list.rds")

```

Looking above, we have created a reusable function for pulling census data from the census API based on the desired year and the desired set of variables. If we wanted to pull a different year, we would update the list of variable names, then simply update the parameters in the new function call to pull that data. We would have to manually include this new data to the census_var_list .rds file we've created at the end, but we could cross that bridge if we came to it.

Next, we are going to write a function to query the PUMS API and return a nicely formatted tibble: 

```{r Part 1 - Parsing API Data 1}

library(httr)
library(jsonlite)
library(tibble)

response_to_tibble <- function(response) {
  httr::stop_for_status(response)

  response_text <- httr::content(
    response,
    as = "text",
    encoding = "UTF-8"
  )

  if (!jsonlite::validate(response_text)) {
    stop(
      "Census returned non-JSON content. Inspect response_text and check your API key.",
      call. = FALSE
    )
  }

  parsed_data <- jsonlite::fromJSON(response_text)

  column_names <- parsed_data[1, ]

  census_data <- tibble::as_tibble(
    parsed_data[-1, , drop = FALSE],
    .name_repair = "minimal"
  )

  names(census_data) <- column_names

  census_data[, !duplicated(names(census_data)), drop = FALSE]
}

api_key <- trimws(Sys.getenv("CENSUS_API_KEY"))

if (!nzchar(api_key)) {
  stop("CENSUS_API_KEY is not set in this R session.")
}

response <- httr::GET(
  "https://api.census.gov/data/2024/acs/acs1/pums",
  query = list(
    get = "AGEP,PWGTP,SEX",
    `for` = "state:37",
    AGEP = "25",
    key = api_key
  )
)

census_data <- response_to_tibble(response)

head(census_data)
str(census_data)


```


Next, since the API returns many values as codes, we need the variable descriptions to know what they all are. We'll write some functions below to extract the codes and labels from the metadata and standardize the codes before matching them:

```{r Part 1 - Parsing API Data 2}

# Return a vector where the names are codes and values are descriptions
get_labels <- function(variable_metadata) {
  labels <- unlist(
    variable_metadata$values$item,
    use.names = TRUE
  )

  if (length(labels) == 0L) {
    return(setNames(character(), character()))
  }

  labels
}

# Standardize numeric codes without losing zero
normalize_code <- function(x) {
  sub("^0+(?=[0-9])", "", as.character(x), perl = TRUE)
}

# Match observed codes to their descriptions
lookup_labels <- function(x, labels) {
  positions <- match(
    normalize_code(x),
    normalize_code(names(labels))
  )

  unname(labels[positions])
}

```


The JWAP and JWDP variables denote the arrival and departure times, but their values are time intervals and not actual clock times. We can use the midpoint of each interval as an approximate time. These functions convert the interval endpoints into seconds after midnight, calculate the midpoint, and store the result as a time of day.

```{r Part 1 - Parsing API Data 3}

# Convert a clock label, such as "4:10 a.m.", to seconds after midnight
clock_seconds <- function(x) {
  cleaned <- tolower(gsub(".", "", trimws(x), fixed = TRUE))

  pieces <- regmatches(
    cleaned,
    regexec("^([0-9]{1,2}):([0-9]{2})\\s*([ap]m)$", cleaned)
  )[[1]]

  if (length(pieces) != 4L) {
    stop("Unrecognized clock label: ", x)
  }

  hour <- as.numeric(pieces[2])
  minute <- as.numeric(pieces[3])

  if (hour < 1 || hour > 12 || minute > 59) {
    stop("Invalid clock label: ", x)
  }

  hour <- hour %% 12

  if (pieces[4] == "pm") {
    hour <- hour + 12
  }

  hour * 3600 + minute * 60
}

# Convert one interval description to its midpoint
interval_midpoint <- function(label) {
  if (is.na(label) ||
      grepl("N/A|not applicable|suppressed", label,
            ignore.case = TRUE)) {
    return(NA_real_)
  }

  endpoints <- strsplit(label, " to ", fixed = TRUE)[[1]]

  if (length(endpoints) != 2L) {
    stop("Unrecognized time interval: ", label)
  }

  start <- clock_seconds(endpoints[1])
  end <- clock_seconds(endpoints[2])

  # Allow an interval to cross midnight
  if (end < start) {
    end <- end + 24 * 3600
  }

  ((start + end) / 2) %% (24 * 3600)
}

# Decode the metadata once, then match the observed codes
parse_census_time <- function(x, variable_metadata) {
  labels <- get_labels(variable_metadata)

  midpoints <- vapply(
    unname(labels),
    interval_midpoint,
    numeric(1)
  )

  positions <- match(
    normalize_code(x),
    normalize_code(names(labels))
  )

  hms::hms(seconds = unname(midpoints[positions]))
}

```

Next, we need to give the remaining variables appropriate types: Numeric values to Numeric type, and categorical values to Factors. Some numeric values represent NA observations and are not actual measurements, and these can be identified using the metadata:

```{r Part 1 - Parsing API Data 4}

parse_census_numeric <- function(x, variable_metadata) {
  labels <- get_labels(variable_metadata)

  missing_codes <- names(labels)[
    grepl("N/A|not applicable|suppressed", labels,
          ignore.case = TRUE)
  ]

  x[normalize_code(x) %in% normalize_code(missing_codes)] <- NA_character_

  result <- suppressWarnings(as.numeric(x))

  # Recognize textual null values, but catch unexpected text
  unexpected <- !is.na(x) &
    !tolower(trimws(x)) %in% c("", "null", "na", "nan") &
    is.na(result)

  if (any(unexpected)) {
    stop("Unexpected nonnumeric values in a numeric variable.")
  }

  result
}

parse_census_factor <- function(x, variable_metadata) {
  labels <- get_labels(variable_metadata)

  if (length(labels) == 0L) {
    stop("No categorical labels found in the metadata.")
  }

  labels <- labels[order(as.numeric(names(labels)))]

  unknown <- !is.na(x) &
    !normalize_code(x) %in% normalize_code(names(labels))

  if (any(unknown)) {
    stop("A categorical code was not found in the metadata.")
  }

  factor(
    normalize_code(x),
    levels = normalize_code(names(labels)),
    labels = unname(labels)
  )
}

```


Now, we'll combine all of our functions above to create the main function query_census. The user can choose a survey year, numeric variables, categorical variables, and geography. Before making a request, the function checks that the inputs meet the proper requirements, specifically: requires PWGTP, at least one other numeric variable, and at least one categorical variable.

The default request uses 2024 data for NC, returning age, sex, and weight. Users can also elect a region/division, or request all valid values within a geography type. Geo codes are checked against the metadata, with PR excluded. Lastly, the function converts the returned columns and adds the "census" class, which allows our custom summary and plotting methods to recognize the data.

```{r Part 1 - Parsing API Data 5}

query_census <- function(year = 2024, numeric_vars = c("AGEP", "PWGTP"), categorical_vars = "SEX", geography = "State", geography_values = NULL) {

  # Validate the year
  if (!is.numeric(year) ||
      length(year) != 1L ||
      anyNA(year) ||
      !year %in% 2021:2024) {
    stop("year must be one of 2021, 2022, 2023, or 2024.")
  }

  allowed_numeric <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"
  )

  allowed_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")

  # Validate numeric variable selections
  if (!is.character(numeric_vars) ||
      anyNA(numeric_vars) ||
      !all(numeric_vars %in% allowed_numeric)) {
    stop("numeric_vars contains an unsupported variable.")
  }

  if (!"PWGTP" %in% numeric_vars ||
      length(setdiff(numeric_vars, "PWGTP")) == 0L) {
    stop("Include PWGTP and at least one other numeric variable.")
  }

  # Validate categorical variable selections
  if (!is.character(categorical_vars) ||
      length(categorical_vars) == 0L ||
      anyNA(categorical_vars) ||
      !all(categorical_vars %in% allowed_categorical)) {
    stop("Select at least one supported categorical variable.")
  }

  numeric_vars <- unique(numeric_vars)
  categorical_vars <- unique(categorical_vars)

  # Validate the geography type
  if (!is.character(geography) ||
      length(geography) != 1L ||
      anyNA(geography) ||
      !geography %in% c("Region", "Division", "State")) {
    stop("geography must be exactly Region, Division, or State.")
  }

  # Load the correct year's metadata
  metadata <- readRDS("data/census_var_list.rds")[[as.character(year)]]

  if (is.null(metadata)) {
    stop("Metadata for the requested year was not found.")
  }

  geography_variable <- switch(
    geography,
    Region = "REGION",
    Division = "DIVISION",
    State = if (year <= 2022) "ST" else "STATE"
  )

  geography_labels <- get_labels(metadata[[geography_variable]])

  # Puerto Rico is outside the scope of this project
  geography_labels <- geography_labels[
    !grepl("Puerto Rico", geography_labels, ignore.case = TRUE)
  ]

  valid_codes <- names(geography_labels)

  # Defaults: North Carolina, South Atlantic, or South
  if (is.null(geography_values)) {
    geography_values <- switch(
      geography,
      State = "37",
      Division = "5",
      Region = "3"
    )
  }

  if (!(is.character(geography_values) ||
        is.numeric(geography_values)) ||
      length(geography_values) == 0L ||
      anyNA(geography_values)) {
    stop("Supply valid geography codes or 'all'.")
  }

  # 'all' is a selection within a geography type
  if (length(geography_values) == 1L &&
      tolower(as.character(geography_values)) == "all") {

    selected_codes <- valid_codes

  } else {

    positions <- match(
      normalize_code(geography_values),
      normalize_code(valid_codes)
    )

    if (anyNA(positions)) {
      stop("Invalid geography code for this year and geography type.")
    }

    selected_codes <- unique(valid_codes[positions])
  }

  api_key <- trimws(Sys.getenv("CENSUS_API_KEY"))

  if (!nzchar(api_key)) {
    stop("Set CENSUS_API_KEY in your R session before querying.")
  }

  response <- httr::GET(
    paste0("https://api.census.gov/data/", year, "/acs/acs1/pums"),
    query = list(
      get = paste(c(numeric_vars, categorical_vars), collapse = ","),
      `for` = paste0(
        tolower(geography), ":",
        paste(selected_codes, collapse = ",")
      ),
      key = api_key
    )
  )

  result <- response_to_tibble(response)

  required_columns <- c(
    numeric_vars, categorical_vars, tolower(geography)
  )

  if (!all(required_columns %in% names(result))) {
    stop("The API response is missing requested columns.")
  }

  # Convert ordinary numeric variables and time variables separately
  for (variable in numeric_vars) {
    if (variable %in% c("JWAP", "JWDP")) {
      result[[variable]] <- parse_census_time(
        result[[variable]], metadata[[variable]]
      )
    } else {
      result[[variable]] <- parse_census_numeric(
        result[[variable]], metadata[[variable]]
      )
    }
  }

  for (variable in categorical_vars) {
    result[[variable]] <- parse_census_factor(
      result[[variable]], metadata[[variable]]
    )
  }

  # Give geography codes readable labels
  geography_column <- tolower(geography)

  result[[geography_column]] <- parse_census_factor(
    result[[geography_column]],
    metadata[[geography_variable]]
  )

  class(result) <- c("census", class(result))

  result
}

```


First we'll test the default call of the function, and also make sure that our time conversion works appropriately:

```{r Part 1 - Parsing API Data 6}

#Testing the default call
nc_2024 <- query_census()

head(nc_2024)
str(nc_2024)
class(nc_2024)

nc_times <- query_census(
  numeric_vars = c("AGEP", "PWGTP", "JWAP", "JWDP")
)

head(nc_times[c("JWAP", "JWDP")])

```


Our single-year function handles one survey year at a time. To compare years, we will write a wrapper that calls it once for each requested year and combines the results. Reusing the existing function keeps the validation and conversion steps consistent. We will also add a year column and store it as a factor because it identifies survey groups for our summaries and plots:

```{r Part 1 - Parsing API Data 8}

query_census_years <- function(years = 2021:2024, numeric_vars = c("AGEP", "PWGTP"), categorical_vars = "SEX", geography = "State", geography_values = NULL) {

  if (!is.numeric(years) ||
      length(years) == 0L ||
      anyNA(years) ||
      !all(years %in% 2021:2024)) {
    stop("years must contain values from 2021 through 2024.")
  }

  years <- unique(years)

  yearly_data <- lapply(years, function(current_year) {

    result <- query_census(
      year = current_year,
      numeric_vars = numeric_vars,
      categorical_vars = categorical_vars,
      geography = geography,
      geography_values = geography_values
    )

    # Bind ordinary tibbles, then restore the custom class
    class(result) <- setdiff(class(result), "census")
    result$year <- as.character(current_year)

    result
  })

  result <- dplyr::bind_rows(yearly_data)
  result$year <- factor(result$year, levels = as.character(sort(years)))

  class(result) <- c("census", class(result))

  result
}
```


# Part 2: Writing a Generic Function for Summarizing

Now we will work on creating some generic functions that we can use to explore our census data, specifically adding methods to the existing summary() and plot() generic functions. We will start with 2 smaller helper functions to identify eligible columns and validate weights. Time variables should not be included in the numeric summaries, and weights must be numeric, non-negative, and finite so that weighted calculations are meaningful:

```{r Part 2 - Writing a Generic Function for Summarizing 1}

is_measurement <- function(x) {
  is.numeric(x) && !inherits(x, c("difftime", "POSIXt", "Date"))
}

check_weights <- function(x) {
  if (!"PWGTP" %in% names(x) ||
      !is.numeric(x$PWGTP) ||
      anyNA(x$PWGTP) ||
      any(!is.finite(x$PWGTP)) ||
      any(x$PWGTP < 0)) {
    stop("PWGTP must contain finite, nonnegative numeric weights.")
  }
}

```


Each row has a person's weight, PWGTP, indicating how much it contributes to the population estimates. Treating every row equally would ignore these differences, so our summary method calculates weighted means and standard deviations using the given formulas, and categorical counts by adding the weights within each category.

By default, the method summarizes all eligible numeric and categorical columns, excluding PWGTP and time variables from the numeric calculations. For each numeric variable, observations with missing values are excluded along with their weights. The results are returned as a named list, separating numeric summaries from categorical counts.

```{r Part 2 - Writing a Generic Function for Summarizing 2}

summary.census <- function(
    object,
    numeric_vars = NULL,
    categorical_vars = NULL,
    ...
) {

  check_weights(object)

  available_numeric <- setdiff(
    names(object)[vapply(object, is_measurement, logical(1))],
    c("PWGTP", "year")
  )

  available_categorical <- names(object)[
    vapply(
      object,
      function(x) is.factor(x) || is.character(x),
      logical(1)
    )
  ]

  if (is.null(numeric_vars)) {
    numeric_vars <- available_numeric
  }

  if (is.null(categorical_vars)) {
    categorical_vars <- available_categorical
  }

  if (!is.character(numeric_vars) ||
      anyNA(numeric_vars) ||
      !all(numeric_vars %in% available_numeric)) {
    stop("Select eligible numeric variables; exclude PWGTP and times.")
  }

  if (!is.character(categorical_vars) ||
      anyNA(categorical_vars) ||
      !all(categorical_vars %in% available_categorical)) {
    stop("Select categorical columns present in the data.")
  }

  numeric_vars <- unique(numeric_vars)
  categorical_vars <- unique(categorical_vars)

  numeric_results <- lapply(numeric_vars, function(variable) {
    values <- object[[variable]]
    weights <- object$PWGTP

    # Use the same observations in numerator and denominator
    keep <- is.finite(values) & weights > 0

    values <- values[keep]
    weights <- weights[keep]

    if (length(values) == 0L) {
      return(c(mean = NA_real_, sd = NA_real_))
    }

    weighted_mean <- sum(values * weights) / sum(weights)

    weighted_variance <- (
      sum(values^2 * weights) / sum(weights)
    ) - weighted_mean^2

    c(
      mean = weighted_mean,
      sd = sqrt(max(0, weighted_variance))
    )
  })

  names(numeric_results) <- numeric_vars

  categorical_results <- lapply(categorical_vars, function(variable) {
    values <- object[[variable]]

    if (!is.factor(values)) {
      values <- factor(values)
    }

    counts <- vapply(levels(values), function(level) {
      sum(object$PWGTP[!is.na(values) & values == level])
    }, numeric(1))

    # Report actual missing observations separately
    if (anyNA(values)) {
      counts <- c(
        counts,
        "(Missing)" = sum(object$PWGTP[is.na(values)])
      )
    }

    counts
  })

  names(categorical_results) <- categorical_vars

  list(
    numeric = numeric_results,
    categorical = categorical_results
  )
}
```


To visually compare, we will create a plotting method for objects with the "census" class. We must select one categorical variable and one numeric measurement. The method produces a boxplot for each category and includes PWGTP in the weight aesthetic so that the boxplot calculations account for the person weights. Observations missing either selected variable, or having zero weight, are excluded from the plot.

```{r Part 2 - Writing a Generic Function for Summarizing 3}
plot.census <- function(x, cat_var, num_var, ...) {

  check_weights(x)

  if (missing(cat_var) || missing(num_var)) {
    stop("Specify one categorical variable and one numeric variable.")
  }

  if (!is.character(cat_var) ||
      length(cat_var) != 1L ||
      anyNA(cat_var) ||
      !cat_var %in% names(x)) {
    stop("cat_var must name one column in the data.")
  }

  if (!is.character(num_var) ||
      length(num_var) != 1L ||
      anyNA(num_var) ||
      !num_var %in% names(x)) {
    stop("num_var must name one column in the data.")
  }

  if (!(is.factor(x[[cat_var]]) || is.character(x[[cat_var]]))) {
    stop("cat_var must be categorical.")
  }

  if (!is_measurement(x[[num_var]]) ||
      num_var %in% c("PWGTP", "year")) {
    stop("num_var must be a numeric measurement, excluding weights and times.")
  }

  keep <- !is.na(x[[cat_var]]) &
    is.finite(x[[num_var]]) &
    x$PWGTP > 0

  plot_data <- x[keep, ]

  if (nrow(plot_data) == 0L) {
    stop("No complete observations with positive weights to plot.")
  }

  class(plot_data) <- setdiff(class(plot_data), "census")

  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(
      x = .data[[cat_var]],
      y = .data[[num_var]],
      weight = PWGTP
    )
  ) +
    ggplot2::geom_boxplot(...) +
    ggplot2::labs(
      x = cat_var,
      y = num_var,
      title = paste(num_var, "by", cat_var),
      subtitle = "Weighted using PWGTP"
    ) +
    ggplot2::theme_minimal()
}
```


We'll now test the functions to make sure that they are working appropriately. Here is a quick example:

```{r Part 2 - Writing a Generic Function for Summarizing 4}

#This is a test data set to check the summary function on
toy <- tibble::tibble(
  AGEP = c(20, 40),
  PWGTP = c(1, 3),
  SEX = factor(c("Male", "Female"))
)

class(toy) <- c("census", class(toy))

toy_summary <- summary(toy)
toy_summary

#This is a test data set to check the plot function on
toy_plot <- tibble::tibble(
  AGEP = c(20, 30, 40, 50, 20, 30, 40, 50),
  PWGTP = c(8, 1, 1, 1, 1, 1, 1, 8),
  SEX = factor(rep(c("Male", "Female"), each = 4))
)

class(toy_plot) <- c("census", class(toy_plot))

p <- plot(
  toy_plot,
  cat_var = "SEX",
  num_var = "AGEP"
)

print(p)

#These checks stop the run if any of the conditions are not true, which they all should be
stopifnot(
  isTRUE(all.equal(unname(toy_summary$numeric$AGEP["mean"]), 35)), #Error if our test example does not have mean = 35
  isTRUE(all.equal(unname(toy_summary$numeric$AGEP["sd"]), sqrt(75))), #Error if our test example does not have sd = sqrt(75)
  toy_summary$categorical$SEX["Male"] == 1,  #Error if our test example does not have 1 male
  toy_summary$categorical$SEX["Female"] == 3 #Error if our test example does not have 3 females
)

#These checks stop the run if any of these interval midpoints are not calculated correctly
stopifnot(
  interval_midpoint("12:00 a.m. to 12:04 a.m.") == 120,
  interval_midpoint("12:00 p.m. to 12:04 p.m.") == 43320,
  interval_midpoint("4:10 a.m. to 4:19 a.m.") == 15270,
  is.na(interval_midpoint("N/A (not a worker)"))
)

#This function prints error messages (intentionally) to ensure that our query_census function is working the right way
show_error <- function(expression) {
  tryCatch(
    expression,
    error = function(e) conditionMessage(e)
  )
}

show_error(query_census(year = 2020))
show_error(query_census(year = c(2023, 2024)))
show_error(query_census(numeric_vars = "AGEP"))
show_error(query_census(numeric_vars = "PWGTP"))
show_error(query_census(categorical_vars = character()))
show_error(query_census(categorical_vars = "MAR"))
show_error(query_census(geography = "All"))
show_error(query_census(geography = c("State", "Region")))
show_error(query_census(geography_values = "99"))


```


